---
title: "SchoolMoves — Pipeline Test: .bin File"
author: "SchoolMoves Pipeline Validation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = TRUE, message = TRUE,
  fig.width = 10, fig.height = 6
)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# 1. Overview

This notebook runs the full GGIR pipeline (Parts 1-5) on the GENEActiv `.bin`
test file and inspects the output at each stage.

**Input**: `data/raw/bin/TESTfile.bin` (389 KB)

**Note**: The test .bin file is very small (389 KB). GGIR requires a minimum
file size for processing (default ~1 MB). If GGIR skips this file, we still
validate the input preparation and format detection steps, and show how the
pipeline handles the error gracefully. For a full pipeline test with .bin files,
use a larger recording file.

```{r load-packages}
library(yaml)
library(dplyr)
library(ggplot2)

# Source pipeline modules
source("R/pipeline/utils.R")
source("R/pipeline/01_prepare_input.R")
source("R/pipeline/02_run_ggir.R")
source("R/pipeline/03_read_ggir_output.R")
source("R/pipeline/04_label_schedule.R")
source("R/pipeline/05_filter_validity.R")
```

# 2. Prepare Input

```{r prepare-input}
input_dir <- file.path("..", "data", "raw", "bin")
output_dir <- file.path("data", "processed", "test_bin")

params <- read_pipeline_params("config/pipeline_params.yaml")

# Scan and classify input files
input_info <- prepare_input(input_dir, output_dir, overwrite = TRUE)
print(input_info$manifest)
```

# 3. Run GGIR (Parts 1-5)

```{r run-ggir, message=TRUE}
ggir_output_dir <- file.path(output_dir, "ggir")

tryCatch({
  run_ggir(
    input_dir = input_info$ggir_input_dir,
    output_dir = ggir_output_dir,
    params = params,
    schedule_path = "config/school_schedules.yaml",
    n_cores = 1
  )
  cat("GGIR completed successfully.\n")
}, error = function(e) {
  cat("GGIR error:", conditionMessage(e), "\n")
  cat("This may be expected for very small test files.\n")
})
```

# 4. Inspect GGIR Output

## 4.1 Output directory structure

```{r inspect-structure}
if (dir.exists(ggir_output_dir)) {
  all_files <- list.files(ggir_output_dir, recursive = TRUE, full.names = FALSE)
  cat("GGIR output files:\n")
  cat(paste(" ", all_files), sep = "\n")
} else {
  cat("No GGIR output directory found.\n")
}
```

## 4.2 GGIR config.csv (reproducibility record)

```{r read-config}
ggir_data <- tryCatch(
  load_all_ggir_output(ggir_output_dir),
  error = function(e) {
    cat("Could not load GGIR output:", e$message, "\n")
    NULL
  }
)

if (!is.null(ggir_data$config)) {
  cat("GGIR parameters used:\n")
  print(head(ggir_data$config, 20))
}
```

## 4.3 Part 2: Daily Summaries

```{r part2-summary}
if (!is.null(ggir_data$part2_daysummary)) {
  cat("Part 2 day summary dimensions:", dim(ggir_data$part2_daysummary), "\n")
  cat("\nColumn names:\n")
  cat(paste(" ", names(ggir_data$part2_daysummary)), sep = "\n")
  cat("\nFirst rows:\n")
  print(head(ggir_data$part2_daysummary))
} else {
  cat("No Part 2 output available.\n")
}
```

## 4.4 Part 4: Sleep Detection

```{r part4-sleep}
if (!is.null(ggir_data$part4_sleep)) {
  cat("Part 4 sleep summary dimensions:", dim(ggir_data$part4_sleep), "\n")
  cat("\nColumn names:\n")
  cat(paste(" ", names(ggir_data$part4_sleep)), sep = "\n")
  cat("\nFirst rows:\n")
  print(head(ggir_data$part4_sleep))
} else {
  cat("No Part 4 sleep output available.\n")
}
```

## 4.5 Part 5: Time-Use Analysis

```{r part5-timeuse}
if (!is.null(ggir_data$part5_daysummary)) {
  cat("Part 5 day summary dimensions:", dim(ggir_data$part5_daysummary), "\n")
  cat("\nColumn names (first 30):\n")
  cat(paste(" ", head(names(ggir_data$part5_daysummary), 30)), sep = "\n")
  cat("\nFirst rows:\n")
  print(head(ggir_data$part5_daysummary[, 1:min(10, ncol(ggir_data$part5_daysummary))]))
} else {
  cat("No Part 5 output available.\n")
}
```

# 5. Visualisations

## 5.1 ENMO Distribution (from Part 2)

```{r plot-enmo, fig.cap="ENMO distribution from Part 2 day summary"}
if (!is.null(ggir_data$part2_daysummary)) {
  enmo_cols <- grep("ENMO|enmo|acceleration", names(ggir_data$part2_daysummary),
                    value = TRUE, ignore.case = TRUE)
  if (length(enmo_cols) > 0) {
    enmo_vals <- as.numeric(ggir_data$part2_daysummary[[enmo_cols[1]]])
    enmo_vals <- enmo_vals[!is.na(enmo_vals)]
    if (length(enmo_vals) > 0) {
      hist(enmo_vals, breaks = 30, main = "ENMO Distribution (Part 2)",
           xlab = "ENMO (mg)", col = "steelblue", border = "white")
      abline(v = c(56.3, 191.6, 695.8), col = c("green", "orange", "red"),
             lwd = 2, lty = 2)
      legend("topright", legend = c("SB/LPA (56.3)", "LPA/MPA (191.6)", "MPA/VPA (695.8)"),
             col = c("green", "orange", "red"), lwd = 2, lty = 2)
    }
  } else {
    cat("No ENMO columns found in Part 2 output.\n")
  }
}
```

## 5.2 Validity Summary

```{r validity}
if (!is.null(ggir_data$part2_daysummary)) {
  validity <- tryCatch(
    assess_validity_from_ggir(ggir_data$part2_daysummary, ggir_data$part4_sleep, params),
    error = function(e) {
      cat("Validity assessment error:", e$message, "\n")
      NULL
    }
  )
  if (!is.null(validity)) {
    print(validity)
  }
}
```

# 6. Summary

```{r summary}
cat("=== .bin Pipeline Test Summary ===\n")
cat("Input file: TESTfile.bin\n")
cat("GGIR output available:", !is.null(ggir_data), "\n")
if (!is.null(ggir_data)) {
  cat("Part 2 (daily summary):", !is.null(ggir_data$part2_daysummary), "\n")
  cat("Part 4 (sleep):", !is.null(ggir_data$part4_sleep), "\n")
  cat("Part 5 (time-use):", !is.null(ggir_data$part5_daysummary), "\n")
  cat("Epoch time series:", !is.null(ggir_data$part5_timeseries), "\n")
}
```
