---
title: "SchoolMoves — Pipeline Test: CSV Files (Dummy Data)"
author: "SchoolMoves Pipeline Validation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = TRUE, message = TRUE,
  fig.width = 10, fig.height = 6
)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# 1. Overview

This notebook runs the full GGIR pipeline on the synthetic GENEActiv CSV files
in `data/example/dummy_data/meting_1/` (10 files, 6 schools, ~40-50 MB each).

These are GENEActiv-format CSVs with 100-row metadata headers, so GGIR can
read them natively.

```{r load-packages}
library(yaml)
library(dplyr)
library(ggplot2)

source("R/pipeline/utils.R")
source("R/pipeline/01_prepare_input.R")
source("R/pipeline/02_run_ggir.R")
source("R/pipeline/03_read_ggir_output.R")
source("R/pipeline/04_label_schedule.R")
source("R/pipeline/05_filter_validity.R")
source("R/pipeline/06_export.R")
source("R/analysis/activity_totals.R")
source("R/analysis/sleep_analysis.R")
```

# 2. Input Inspection

```{r inspect-input}
input_dir <- file.path("..", "data", "example", "dummy_data", "meting_1")

csv_files <- list.files(input_dir, pattern = "\\.csv$", full.names = TRUE)
cat("Number of CSV files:", length(csv_files), "\n")
cat("Files:\n")
for (f in csv_files) {
  cat(sprintf("  %s (%.1f MB)\n", basename(f), file.size(f) / 1e6))
}

# Read one file header to confirm format
if (length(csv_files) > 0) {
  header <- readLines(csv_files[1], n = 25)
  cat("\nFirst 5 header lines of", basename(csv_files[1]), ":\n")
  cat(paste(header[1:5], collapse = "\n"), "\n")

  subject_line <- header[21]
  cat("\nSubject Code line:", subject_line, "\n")
}
```

# 3. Prepare Input

```{r prepare-input}
output_dir <- file.path("data", "processed", "test_csv")
params <- read_pipeline_params("config/pipeline_params.yaml")

input_info <- prepare_input(input_dir, output_dir, overwrite = TRUE)
print(input_info$manifest[, c("filename", "format", "pupil_id", "school_id", "ggir_compatible")])
```

# 4. Run GGIR (Parts 1-5)

```{r run-ggir, message=TRUE}
ggir_output_dir <- file.path(output_dir, "ggir")

t_start <- Sys.time()
tryCatch({
  run_ggir(
    input_dir = input_info$ggir_input_dir,
    output_dir = ggir_output_dir,
    params = params,
    schedule_path = "config/school_schedules.yaml",
    n_cores = 1
  )
  cat("GGIR completed in", round(difftime(Sys.time(), t_start, units = "mins"), 1), "minutes\n")
}, error = function(e) {
  cat("GGIR error:", conditionMessage(e), "\n")
})
```

# 5. Inspect GGIR Output

## 5.1 Output Directory

```{r output-structure}
if (dir.exists(ggir_output_dir)) {
  all_files <- list.files(ggir_output_dir, recursive = TRUE, full.names = FALSE)
  cat("Total output files:", length(all_files), "\n\n")

  # Group by directory
  dirs <- unique(dirname(all_files))
  for (d in sort(dirs)) {
    files_in_dir <- all_files[dirname(all_files) == d]
    cat(d, "/\n")
    cat(paste("  ", basename(files_in_dir)), sep = "\n")
    cat("\n")
  }
}
```

## 5.2 Load All Output

```{r load-output}
ggir_data <- tryCatch(
  load_all_ggir_output(ggir_output_dir),
  error = function(e) {
    cat("Could not load GGIR output:", e$message, "\n")
    NULL
  }
)
```

## 5.3 Part 2: Daily Summaries

```{r part2}
if (!is.null(ggir_data$part2_daysummary)) {
  p2 <- ggir_data$part2_daysummary
  cat("Part 2 dimensions:", nrow(p2), "rows x", ncol(p2), "columns\n")
  cat("\nColumn names:\n")
  cat(paste(" ", names(p2)), sep = "\n")
  cat("\n\nFirst 6 rows (first 8 columns):\n")
  print(head(p2[, 1:min(8, ncol(p2))]))
}
```

## 5.4 Part 4: Sleep

```{r part4}
if (!is.null(ggir_data$part4_sleep)) {
  p4 <- ggir_data$part4_sleep
  cat("Part 4 dimensions:", nrow(p4), "rows x", ncol(p4), "columns\n")
  cat("\nFirst 6 rows (first 8 columns):\n")
  print(head(p4[, 1:min(8, ncol(p4))]))
}
```

## 5.5 Part 5: Time-Use

```{r part5}
if (!is.null(ggir_data$part5_daysummary)) {
  p5 <- ggir_data$part5_daysummary
  cat("Part 5 dimensions:", nrow(p5), "rows x", ncol(p5), "columns\n")
  cat("\nFirst 20 column names:\n")
  cat(paste(" ", head(names(p5), 20)), sep = "\n")
}
```

# 6. School Context Labelling

```{r context-labelling}
qwindow <- build_qwindow_from_schedules("config/school_schedules.yaml")
context_map <- build_qwindow_context_map("config/school_schedules.yaml", qwindow)

cat("qwindow boundaries (decimal hours):\n")
print(qwindow)

cat("\nContext map (school 1):\n")
print(context_map[context_map$school_id == 1, ])
```

# 7. Validity Assessment

```{r validity}
if (!is.null(ggir_data$part2_daysummary)) {
  validity <- tryCatch(
    assess_validity_from_ggir(ggir_data$part2_daysummary, ggir_data$part4_sleep, params),
    error = function(e) {
      cat("Validity error:", e$message, "\n")
      NULL
    }
  )

  if (!is.null(validity)) {
    cat("\nValidity summary:\n")
    print(validity)

    cat("\n\nValid for waking analysis:", sum(validity$pupil_valid_sedentary),
        "/", nrow(validity), "\n")
    cat("Valid for sleep analysis:", sum(validity$pupil_valid_sleep),
        "/", nrow(validity), "\n")
  }
}
```

# 8. Activity Totals (GGIR-based)

```{r activity-totals}
if (!is.null(ggir_data$part5_daysummary)) {
  activity <- compute_activity_totals_ggir(ggir_data$part5_daysummary, context_map)
  if (!is.null(activity)) {
    cat("Activity totals dimensions:", nrow(activity), "rows x", ncol(activity), "columns\n")
    cat("\nPupil IDs found:", paste(unique(activity$pupil_id), collapse = ", "), "\n")
    cat("School IDs found:", paste(unique(activity$school_id), collapse = ", "), "\n")
  }
}
```

# 9. Sleep Analysis (GGIR-based)

```{r sleep-analysis}
if (!is.null(ggir_data$part4_sleep)) {
  sleep <- summarise_sleep_ggir(ggir_data$part4_sleep)
  if (nrow(sleep) > 0) {
    cat("Sleep summary dimensions:", nrow(sleep), "rows\n")
    print(head(sleep))
  }
}
```

# 10. Visualisations

## 10.1 Per-School Summary

```{r per-school-plot, fig.cap="GGIR output summary per school"}
if (!is.null(ggir_data$part2_daysummary)) {
  p2 <- ggir_data$part2_daysummary
  # Try to find useful numeric columns
  num_cols <- names(p2)[sapply(p2, is.numeric)]
  if (length(num_cols) > 2) {
    cat("Available numeric columns:", paste(head(num_cols, 10), collapse = ", "), "\n")
  }
}
```

## 10.2 Pupils per School

```{r pupils-per-school, fig.cap="Number of pupils per school in input"}
manifest <- input_info$manifest
if (nrow(manifest) > 0) {
  school_counts <- table(manifest$school_id)
  barplot(school_counts, main = "Pupils per School (Input Files)",
          xlab = "School ID", ylab = "Number of Files",
          col = "steelblue", border = "white")
}
```

# 11. Summary

```{r final-summary}
cat("=== CSV Pipeline Test Summary ===\n")
cat("Input files:", nrow(input_info$manifest), "\n")
cat("GGIR-compatible:", sum(input_info$manifest$ggir_compatible), "\n")
cat("Schools:", paste(sort(unique(input_info$manifest$school_id)), collapse = ", "), "\n")
cat("\nGGIR output:\n")
if (!is.null(ggir_data)) {
  cat("  Part 2 (daily):", !is.null(ggir_data$part2_daysummary),
      if (!is.null(ggir_data$part2_daysummary)) paste0("(", nrow(ggir_data$part2_daysummary), " rows)"), "\n")
  cat("  Part 4 (sleep):", !is.null(ggir_data$part4_sleep),
      if (!is.null(ggir_data$part4_sleep)) paste0("(", nrow(ggir_data$part4_sleep), " rows)"), "\n")
  cat("  Part 5 (time-use):", !is.null(ggir_data$part5_daysummary),
      if (!is.null(ggir_data$part5_daysummary)) paste0("(", nrow(ggir_data$part5_daysummary), " rows)"), "\n")
}
```
