---
title: "SchoolMoves — Comparison: Old Program vs New GGIR Pipeline"
author: "SchoolMoves Pipeline Validation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = TRUE, message = TRUE,
  fig.width = 10, fig.height = 6
)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# 1. Purpose

This notebook investigates **where** and **why** the old program
(`data/raw/oud_programma/1_general.R`) and the new GGIR-based pipeline differ.

The goal is NOT to force alignment but to document and understand the
methodological differences, identify expected vs unexpected divergences,
and assess whether the new pipeline's outputs are scientifically valid.

```{r load-packages}
library(dplyr)
library(ggplot2)
library(yaml)

source("R/pipeline/utils.R")
source("R/pipeline/03_read_ggir_output.R")
source("R/pipeline/04_label_schedule.R")
```

# 2. Methodology Comparison

## 2.1 High-Level Differences

```{r methodology-table}
comparison <- data.frame(
  Dimension = c(
    "Device", "Body location", "GGIR usage",
    "Activity classification", "Epoch length",
    "Sleep detection", "Non-wear detection",
    "Bout detection", "School context source",
    "Output granularity"
  ),
  Old_Program = c(
    "Axivity", "Thigh", "Part 1 only (ENMO + nonwear)",
    "Random Forest -> activity TYPE (Sit/Stand/Walk/Run/Cycle/Lie)",
    "5 seconds", "External Axivity file + SPSS manual records",
    "GGIR Part 1 + student-reported (from SPSS)",
    "Custom sitting bouts, 30-sec interruption rule, 6 duration bins",
    "SPSS .sav files (7 lessons per day)",
    "5-second epochs, day-level aggregation, bout-level"
  ),
  New_Pipeline = c(
    "GENEActiv", "Wrist", "Parts 1-5 (full pipeline)",
    "ENMO cut-points -> intensity LEVEL (SB/LPA/MPA/VPA)",
    "5 seconds (GGIR default windowsize)", "GGIR Part 4 HDCZA algorithm",
    "GGIR Part 2 integrated (SD + range + duration)",
    "GGIR boutdur params + custom context-split bouts",
    "config.yaml (optional .sav support)",
    "GGIR day-segment summaries, person summaries"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(comparison, col.names = c("Dimension", "Old Program", "New Pipeline"),
             caption = "Methodological comparison")
```

## 2.2 Why These Differences Exist

### Device & Body Location

The old program was designed for an **Axivity thigh-worn** accelerometer study.
The current project uses **GENEActiv wrist-worn** devices. This is the most
fundamental difference: wrist-worn acceleration patterns are qualitatively
different from thigh-worn (more arm movement artifacts, different non-wear
signatures, different cut-point thresholds).

**Implication**: Cut-points, activity classification, and sleep detection
algorithms are NOT interchangeable between body locations.

### Activity Classification Approach

| Old (Thigh) | New (Wrist) |
|---|---|
| Random Forest model | ENMO cut-points (Hildebrand 2014/2017) |
| 6 activity TYPES: Sitting, Lying, Standing, Walking, Running, Cycling | 4 intensity LEVELS: SB, LPA, MPA, VPA |
| Requires training data and pre-built model | Uses validated published thresholds |
| Position-specific features (thigh angle) | Acceleration magnitude only |

**Why different**: Thigh-worn devices can distinguish postures (sitting vs
standing) via inclination angle, enabling activity TYPE classification.
Wrist-worn devices cannot reliably distinguish postures, so intensity-based
classification using ENMO thresholds is the standard approach.

### GGIR Usage

The old program used GGIR Part 1 **only** for ENMO extraction and nonwear
scoring, then did everything else with custom code. The new pipeline uses GGIR
Parts 1-5, getting validated implementations of:

- Autocalibration (Part 1)
- Non-wear detection (Part 2)
- Cut-point classification (Part 2)
- Sleep detection (Parts 3-4)
- Time-use analysis with bout detection (Part 5)

# 3. Output Structure Comparison

## 3.1 Old Program Outputs

```{r old-output-structure}
old_outputs <- data.frame(
  File = c("PA_per_5sec.csv", "activity_per_schoolday.Rds", "PA_per_bout.Rds"),
  Level = c("5-second epoch", "Day-level", "Bout-level"),
  Key_Columns = c(
    "timestamp, pid, activity_type, ENMO, during_class_1..7, sitting, bout_nummer_sitting, nonwear_*",
    "Sitting_total, Standing_total, Walking_total (+ per context), ENMO means, bout counts by duration bin",
    "bout_nummer_sitting, bout_time_sitting, bout_time_sitting_exclusive_school/in_class/out_class"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(old_outputs, caption = "Old program output files")
```

## 3.2 New Pipeline Outputs

```{r new-output-structure}
new_outputs <- data.frame(
  File = c(
    "part2_daysummary.csv", "part4_nightsummary_sleep_cleaned.csv",
    "part5_daysummary_*.csv", "part5_personsummary_*.csv",
    "validity_report.csv", "analysis/activity_totals.csv",
    "analysis/sleep_summary.csv"
  ),
  Level = c("Day-level", "Night-level", "Day-segment-level",
            "Person-level", "Person-level", "Day-context-level", "Night-level"),
  Source = c("GGIR Part 2", "GGIR Part 4", "GGIR Part 5",
             "GGIR Part 5", "Custom", "Custom (from GGIR)", "Custom (from GGIR)"),
  stringsAsFactors = FALSE
)
knitr::kable(new_outputs, caption = "New pipeline output files")
```

## 3.3 Column Mapping

```{r column-mapping}
col_map <- data.frame(
  Concept = c(
    "Activity intensity", "School context", "Non-wear flag",
    "Sleep period", "Bout duration", "ENMO metric",
    "Observation period", "Absence"
  ),
  Old_Column = c(
    "activity_type (Sitting/Lying/Standing/Walking/Running/Cycling)",
    "during_class_1..7, at_school_activity, in_class_activity",
    "nonwear_detected, nonwear_reported, nonwear_unique",
    "sleep_axivity (from SPSS)",
    "bout_time_sitting, bout_time_sitting_exclusive_*",
    "ENMO (from GGIR Part 1, per minute)",
    "during_observation (from SPSS admin)",
    "absent (from SPSS admin)"
  ),
  New_Column = c(
    "GGIR Part 2 intensity flags (dur_day_IN_*, dur_day_LIG_*, etc.)",
    "qwindow segments mapped to context via config.yaml",
    "GGIR Part 2 integrated non-wear (N valid hours)",
    "GGIR Part 4 sleeponset/wakeup",
    "GGIR Part 5 boutdur.in/boutdur.mvpa columns",
    "GGIR Part 2 ENMO (per epoch, calibrated)",
    "Full recording period (no explicit observation window)",
    "Not available (requires .sav metadata)"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(col_map, caption = "Column mapping between old and new")
```

# 4. Data-Level Comparison

This section loads actual pipeline output (if available) and compares
distributions and values.

## 4.1 Load New Pipeline Output

```{r load-new-output}
# Try loading from the CSV test run
csv_output_dir <- file.path("data", "processed", "test_csv", "ggir")
bin_output_dir <- file.path("data", "processed", "test_bin", "ggir")

new_data <- list()

for (dir_name in c("csv" = csv_output_dir, "bin" = bin_output_dir)) {
  label <- names(which(c(csv_output_dir, bin_output_dir) == dir_name))
  tryCatch({
    data <- load_all_ggir_output(dir_name)
    new_data[[label]] <- data
    cat("Loaded", label, "output\n")
  }, error = function(e) {
    cat("Could not load", label, "output:", e$message, "\n")
  })
}
```

## 4.2 Read Old Program Code Structure

```{r parse-old-program}
old_script_path <- file.path("..", "data", "raw", "oud_programma", "1_general.R")

if (file.exists(old_script_path)) {
  old_code <- readLines(old_script_path, warn = FALSE)
  cat("Old program: ", length(old_code), "lines of R code\n\n")

  # Extract key function calls
  ggir_calls <- grep("GGIR|g\\.part", old_code, value = TRUE)
  cat("GGIR-related lines:\n")
  cat(paste("  ", trimws(ggir_calls)), sep = "\n")

  # Extract data sources
  data_sources <- grep("read_rds|read_spss|read_csv", old_code, value = TRUE)
  cat("\n\nData source lines:\n")
  cat(paste("  ", trimws(head(data_sources, 10))), sep = "\n")
} else {
  cat("Old program script not found at:", old_script_path, "\n")
}
```

# 5. Key Difference Analysis

## 5.1 Activity Classification: TYPE vs INTENSITY

```{r classification-comparison}
cat("=== Activity Classification Comparison ===\n\n")
cat("OLD PROGRAM (Thigh-worn, Random Forest):\n")
cat("  Classes: Sitting, Lying, Standing, Walking, Running, Cycling\n")
cat("  Method: Random Forest model trained on labelled thigh-accelerometer data\n")
cat("  Granularity: Activity TYPE (what are you doing?)\n")
cat("  Epoch: 5 seconds\n\n")

cat("NEW PIPELINE (Wrist-worn, Cut-points):\n")
cat("  Classes: Sedentary (<56.3 mg), Light (56.3-191.6), Moderate (191.6-695.8), Vigorous (>695.8)\n")
cat("  Method: Hildebrand et al. 2014/2017 wrist-worn children cut-points\n")
cat("  Granularity: Activity INTENSITY (how hard are you moving?)\n")
cat("  Epoch: 5 seconds (GGIR default)\n\n")

cat("EXPECTED DIFFERENCE: These are fundamentally different classification\n")
cat("approaches. They cannot be directly compared numerically.\n")
cat("However, we can compare the PROPORTIONS:\n")
cat("  Old 'Sitting + Lying' ~ New 'Sedentary'\n")
cat("  Old 'Standing' ~ New 'Light' (approximately)\n")
cat("  Old 'Walking' ~ New 'Light/Moderate'\n")
cat("  Old 'Running + Cycling' ~ New 'Moderate/Vigorous'\n")
```

## 5.2 Bout Detection

```{r bout-comparison}
cat("=== Bout Detection Comparison ===\n\n")

cat("OLD PROGRAM:\n")
cat("  Target: Sitting/Lying bouts (posture-based)\n")
cat("  Interruption rule: Non-sitting episode < 30 seconds = part of sitting bout\n")
cat("  Duration bins: 0-1, 1-4, 5-9, 10-19, 20-29, 30+ minutes\n")
cat("  Context split: Bouts split at school context boundaries\n")
cat("  Context types: at_school, in_class, out_class, elsewhere\n\n")

cat("NEW PIPELINE:\n")
cat("  Target: Inactivity bouts (ENMO-based, <56.3 mg)\n")
cat("  Interruption rule: GGIR's boutcriter parameter (default 80% of bout)\n")
cat("  Duration bins: 10, 20, 30 minutes (configurable via boutdur.in)\n")
cat("  Context split: Custom context-split bouts available (our value-add)\n")
cat("  Context types: in_class, recess, lunch, before_school, after_school, weekend\n\n")

cat("KEY DIFFERENCE: Old program detects SITTING bouts (posture), new pipeline\n")
cat("detects SEDENTARY bouts (low acceleration). A child standing still would\n")
cat("be 'Standing' in old program but 'Sedentary' in new pipeline.\n")
```

## 5.3 Sleep Detection

```{r sleep-comparison}
cat("=== Sleep Detection Comparison ===\n\n")

cat("OLD PROGRAM:\n")
cat("  Source: Axivity device sleep detection (external)\n")
cat("  Stored in: SPSS .sav file (Databestand_Slaap.sav)\n")
cat("  Method: Likely angle-based detection from thigh accelerometer\n")
cat("  Override: Sets activity_type = 'Sleeping' during detected sleep\n\n")

cat("NEW PIPELINE:\n")
cat("  Source: GGIR Part 4 (HDCZA algorithm)\n")
cat("  Method: Heuristic algorithm using sustained inactivity bouts (Part 3)\n")
cat("    + arm angle change detection from wrist accelerometer\n")
cat("  Output: sleeponset, wakeup, SleepDuration per night\n\n")

cat("KEY DIFFERENCE: Different devices, different body locations, different\n")
cat("algorithms. Wrist-based sleep detection (HDCZA) is well-validated for\n")
cat("population-level studies. Exact onset/offset times may differ by 15-30 min.\n")
```

## 5.4 Non-Wear Detection

```{r nonwear-comparison}
cat("=== Non-Wear Detection Comparison ===\n\n")

cat("OLD PROGRAM:\n")
cat("  Sources (3):\n")
cat("    1. GGIR Part 1 nonwear score (automated, per minute)\n")
cat("    2. Student-reported non-wear times (from SPSS Dataset_Leerlingen.sav)\n")
cat("    3. Union of both: nonwear_unique = max(detected, reported)\n\n")

cat("NEW PIPELINE:\n")
cat("  Source: GGIR Part 2 integrated non-wear detection\n")
cat("  Method: SD < threshold AND range < threshold on >= 2 of 3 axes\n")
cat("    over 60-min blocks assessed every 15 min\n")
cat("  No student-reported data (unless .sav metadata provided)\n\n")

cat("KEY DIFFERENCE: Old program had richer non-wear data (student self-report).\n")
cat("New pipeline relies solely on algorithmic detection unless .sav files are\n")
cat("provided. This may slightly overestimate wear time.\n")
```

# 6. Comparison Visualisations

## 6.1 Pipeline Flow Comparison

```{r pipeline-flow, fig.height=8, fig.cap="Pipeline flow comparison"}
# Text-based flow diagram
cat("
OLD PROGRAM FLOW:                    NEW PIPELINE FLOW:
========================             ========================
Raw .cwa files (Axivity)             Raw .bin/.csv (GENEActiv)
         |                                    |
  GGIR g.part1() ONLY               GGIR Parts 1-5
  (ENMO + nonwear)                   (calibration, ENMO, nonwear,
         |                            sleep, time-use, bouts)
  Random Forest features                      |
  (5-sec epoch features)             Part 2: daily summaries
         |                           Part 4: sleep per night
  RF Activity Type prediction        Part 5: time-use per segment
  (Sit/Stand/Walk/Run/Cycle)                  |
         |                           School context labelling
  SPSS admin data merge              (config.yaml qwindow mapping)
  (schedules, absences, sleep)                |
         |                           Validity assessment
  Custom bout detection              (GGIR validity flags)
  (sitting bouts, 30-sec rule)                |
         |                           Export
  Day-level aggregation              (GGIR CSVs + analysis CSVs)
  (activity min per context)
         |
  Export (Rds, sav, csv)
")
```

## 6.2 School Context Resolution

```{r context-resolution}
# Compare school context approaches
schedules <- load_school_schedules("config/school_schedules.yaml")

cat("=== School Context Resolution ===\n\n")
cat("OLD PROGRAM: 7 individual lesson flags (during_class_1 .. during_class_7)\n")
cat("  Each lesson has explicit start/end from SPSS admin data\n")
cat("  Combined into: at_school, in_class, out_class, out_school\n\n")

cat("NEW PIPELINE: Named context blocks from config.yaml\n")
for (sch in schedules[1:2]) {
  cat(sprintf("\n  School %d (%s):\n", sch$school_id, sch$schedule_source))
  for (block in sch$blocks) {
    cat(sprintf("    %s - %s: %s\n", block$start, block$end, block$context))
  }
}

cat("\n\nMapping between old and new:\n")
cat("  OLD 'during_class_1..7'  -> NEW 'in_class'\n")
cat("  OLD 'during_school_hours & !during_class'  -> NEW 'recess' or 'lunch'\n")
cat("  OLD '!during_school_hours'  -> NEW 'before_school' or 'after_school'\n")
cat("  OLD 'weekend flag (via wday)'  -> NEW 'weekend'\n")
```

# 7. Expected vs Unexpected Differences

```{r expected-differences}
cat("=== EXPECTED DIFFERENCES (methodologically justified) ===\n\n")
cat("1. Activity classification outputs are INCOMPARABLE\n")
cat("   (TYPE vs INTENSITY, thigh vs wrist)\n\n")
cat("2. Bout definitions differ\n")
cat("   (sitting bouts vs sedentary bouts, different interruption rules)\n\n")
cat("3. Sleep onset/offset will differ by 15-30 min\n")
cat("   (different algorithms, different body locations)\n\n")
cat("4. Non-wear estimates may differ\n")
cat("   (old has student self-report, new is algorithmic only)\n\n")
cat("5. ENMO values will differ\n")
cat("   (wrist movement >> thigh movement during sedentary tasks)\n\n")
cat("6. School context granularity differs\n")
cat("   (old: 7 lesson flags, new: named context blocks)\n\n")

cat("\n=== POTENTIALLY UNEXPECTED DIFFERENCES (investigate if found) ===\n\n")
cat("1. Very different wear-time totals for same participant\n")
cat("   -> Check non-wear algorithm sensitivity\n\n")
cat("2. Total school-hours activity time differs by >30 min/day\n")
cat("   -> Check schedule alignment\n\n")
cat("3. No sleep detected for a participant where old program found sleep\n")
cat("   -> Check HDCZA sensitivity parameters\n\n")
cat("4. GGIR fails to process files that old program handled\n")
cat("   -> Check file format compatibility\n")
```

# 8. Recommendations

```{r recommendations}
cat("=== RECOMMENDATIONS FOR INTERPRETATION ===\n\n")

cat("1. DO NOT directly compare intensity minutes between old and new pipelines.\n")
cat("   They measure different things (activity type vs intensity level).\n\n")

cat("2. DO compare structural patterns:\n")
cat("   - Are the same participants flagged as invalid?\n")
cat("   - Do school-hour activity patterns show similar temporal structure?\n")
cat("   - Are sedentary bout counts in the same order of magnitude?\n\n")

cat("3. For publication, report the new pipeline's methods (GGIR Parts 1-5,\n")
cat("   Hildebrand cut-points, HDCZA sleep detection) and cite:\n")
cat("   - van Hees et al. (GGIR methods papers)\n")
cat("   - Hildebrand et al. 2014, 2017 (cut-points)\n\n")

cat("4. Consider running a small subsample through both pipelines on the SAME\n")
cat("   device data to quantify the impact of algorithm choice vs device choice.\n")
```

# 9. Session Info

```{r session-info}
sessionInfo()
```
